Supplementary Information

### 2020-11-07 doc/manuscript/MEE_manuscript_SI.Rmd
deleted figures, now knitting works

```{r eval = TRUE, echo=TRUE, warning=FALSE} 
#, results='hide'}
#READING IN TABLES TO GET STATS FOR TEXT
path_to_dir <- dirname(dirname(getwd())) # get path to repo (ForestGEO-climate-sensitivity)
```

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
breakrow1<-10
breakrow2<-24
breakrow3<-35
col_width<-"2.5cm"
table_pt1 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=0, nrows=breakrow1,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt1, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Barro Colorado Island, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```

@@

\newpage
Table S5, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Lilly Dickey Woods, Indiana, USA", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Harvard Forest, Massachusetts, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Niobrara and Hansley, Nebraska, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  #pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  #pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
---
Results from this study are the climate-only model. 
Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings. 
Beyond the methodological differences, original studies vary from this one and from one another in factors including
exact set of cores analyzed,
climate data sources,
time frame of analysis,
approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
methods for detrending and standardizing to build chronologies,
and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
An analysis standardizing all of these factors for four species is present in Appendix **S2** and Fig. **S1**.

@@

\newpage
## Figure S2. Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA)
```{r , echo=FALSE, fig.align = "center", out.width='75%', fig.pos='H'}

  knitr::include_graphics(paste0(path_to_dir, "/results/climwin_plots_combined/NewMexico_pet.png"))
```

**Figure S2 | Comparison of climwin output across growth metrics for the temperature variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified potential evapotranspiration ($PET$) as the strongest climate variable across all three metrics of growth ($RW$, BAI, $\Delta AGB$) and regardless of whether all cores were included in the analysis, or only those for which DBH could be reconstructed ($RW$-trees with $DBH$, BAI, $\Delta AGB$).

\newpage
## Figure S3. Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA)
![**Figure S3 | Comparison of climwin output across growth metrics for the precipitation variable group at Little Tesuque (New Mexico, USA).** Here, *climwin* identified precipitation ($PPT$) as the strongest climate variable for $RW$ and $BAI$, but precipitation day frequency ($PDF$) as the strongest climate variable for $\Delta AGB$.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/NewMexico_wet_pre.png)

\newpage
## Figure S4. Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA)
![**Figure S4 | Comparison of climwin output across growth metrics for the precipitation variable group at Harvard Forest (Massachusetts, USA).** Here, *climwin* identified precipitation frequency ($PDF$) as the strongest climate variable for $RW$, but precipitation amount ($PPT$) as the strongest climate variable for $BAI$ and $\Delta AGB$. The optimal time window (circled) also differed across growth metrics.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/climwin_plots_combined/HarvardForest_pre_wet.png)

\newpage
## Figure S5. Best GLS models for Barro Colorado Island (Panama)

![**Figure S5 | Best GLS models for Barro Colorado Island (Panama) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/BCI.png){ width=75% }

\newpage
## Figure S6. Best GLS models for Huai Kha Khaeng (Thailand)

![**Figure S6 | Best GLS models for Huai Kha Khaeng (Thailand) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HKK.png){ width=75% }

\newpage
## Figure S7. Best GLS models for Little Tesuque (New Mexico, USA)

![**Figure S7 | Best GLS models for Little Tesuque  (New Mexico, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/NewMexico.png){ width=75% }

\newpage
## Figure S8. Best GLS models for Cedar Breaks (Utah, USA)

[**Figure S8 | Best GLS models for Cedar Breaks  (Utah, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.]

\newpage
## Figure S9. Best GLS models for SCBI (Virginia, USA)

![**Figure S9 | Best GLS models for SCBI (Virginia, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/SCBI.png){ width=75% }

\newpage
## Figure S10. Best GLS models for Lilly Dickey Woods (Indiana, USA)

![**Figure S10 | Best GLS models for Lilly Dickey Woods (Indiana, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/LillyDickey.png){ width=75% }

\newpage
## Figure S11. Best GLS models for Harvard Forest (Massachusetts, USA)

![**Figure S11 | Best GLS models for Harvard Forest (Massachusetts, USA) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/HarvardForest.png){ width=75% }

\newpage
## Figure S12. Best GLS models for Niobrara/ Hansley (Nebraska, USA)

\newpage
## Figure S13. Best GLS models for Zofin (Czech Republic)

![**Figure S13 | Best GLS models for Zofin (Czech Republic) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/Zofin.png){ width=75% }

\newpage
## Figure S14. Best GLS models for Scotty Creek (NW Territories, Canada)

![**Figure S14 | Best GLS models for Scotty Creek (NW Territories, Canada) for all three growth metrics: $RW$, $BAI$, and $\Delta AGB$.** Precipitation and temperature group variables are as selected by *climwin* (p=previous year, c=current year). For each species, relationships are plotted if included in top model, with best-fit polynomials plotted with solid lines when both first- and second-order terms are signficant, dashed lines when only one term is signficant, and dotted lines when neither is signficant. Transparent ribbons indicate 95% confidence intervals. Vertical grey lines indicate the long-term mean for the climate variable, shading indicates 1 SD.](/Users/kteixeira/Dropbox (Smithsonian)/GitHub/EcoClimLab/ForestGEO-climate-sensitivity/results/composite_plots/ScottyCreek.png){ width=75% }

>>>>

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
#READING IN TABLES TO GET STATS FOR TEXT
path_to_dir <- dirname(dirname(getwd())) # get path to repo (ForestGEO-climate-sensitivity)
```

@@

```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
breakrow1<-10
breakrow2<-24
breakrow3<-37
col_width<-"2.5cm"
table_pt1 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=0, nrows=breakrow1,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt1, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Barro Colorado Island, Panama", 1, 3, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Huai Kha Khaeng, Thailand", 4, 7, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```

@@

\newpage
Table S5, cont.
```{r eval = TRUE, echo=FALSE, warning=FALSE}
table_pt3 <- read.csv("tables_figures/qualitative_results_comparison.csv", 
                     colClasses=c("NULL", NA, NA, NA, NA, NA, NA), 
                     col.names=c("NULL", "species", "previously observed", "observed here", "previously observed", "observed here", "reference"), 
                     skip=breakrow2, nrows=breakrow3-breakrow2,
                     stringsAsFactors = FALSE, check.names = FALSE)
kable(table_pt3, format = "latex", booktabs = TRUE, longtable = TRUE,  escape = F, linesep = c('\\addlinespace'))%>%
  add_header_above(c(" ", "Precipitation response" = 2, "Temperature response" = 2)) %>%
  column_spec(2:5, width = col_width) %>%
  column_spec(6, width = "2cm") %>%
  pack_rows("Lilly Dickey Woods, Indiana, USA", 1, 4, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Harvard Forest, Massachusetts, USA", 5, 5, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Niobrara and Hansley, Nebraska, USA", 6, 8, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Žofín Forest Dynamics Plot, Czech Republic", 9, 12, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  pack_rows("Scotty Creek, NW Territories, Canada", 13, 13, latex_gap_space = "1em", colnum = 4, hline_before = FALSE) %>%
  kable_styling(latex_options = c( "hold_position", "repeat header"), protect_latex = T, font_size = 7) 
# refs in this table need to go in the nocite field below
```
---
nocite: | 
  @alfaro-sanchez_growth_2017; @vlam_temperature_2014; @helcoski_growing_2019; @maxwell_declining_2016; @aus_de_ar_tree_2018; @bumann_assessing_2019; @tumajer_increasing_2017; @sniderhan_growth_2016; @kaspar_species-specific_nodate
---
Results from this study are the climate-only model. 
Where previous studies examined numerous climate variables or time windows [e.g., @helcoski_growing_2019], we focus on those most relevant to our findings. 
Beyond the methodological differences, original studies vary from this one and from one another in factors including
exact set of cores analyzed,
climate data sources,
time frame of analysis,
approaches to identifying candidate climate variables and windows (including whether this is done on a site or species level), 
methods for detrending and standardizing to build chronologies,
and whether the effects of temperature and precipitation are considered separately (original studies) or additively (this study).
An analysis standardizing all of these factors for four species is present in Appendix **S2** and Fig. **S1**.

@@
